The Evolution of Phenotypically Invariant Genetic Networks

### 2017-10-03 kalman_walk_paper.tex
edited envelope calc

\documentclass{article}
\usepackage{amsmath, amssymb, color, xcolor, amsthm}
\usepackage{graphicx, wrapfig, float, caption, dsfont, bbm}
\usepackage{fullpage}
\usepackage[backref=page, hidelinks, colorlinks=true, citecolor=blue!60!black!100]{hyperref}
\usepackage{tikz}
\usetikzlibrary{arrows.meta, shapes}
\usepackage{caption, subcaption}
\usepackage{natbib} % gives us \citet: Author (year) and \citep: (Author; year)
\usepackage{authblk}

@@

\begin{figure}
\centering
\includegraphics{figures/conceptual_fig}
\caption{
    \label{fig:conceptual_fig}
    A conceptual figure of the fitness consequences of hybridization:
    axes represent system coefficients (i.e., entries of $A$);
    the line of optimal system coefficients is down in black;
    dotted lines give phenotypic distances to the optimum.
    Two pairs of parental populations are shown in black, along the optimum;
    a hypothetical population of $F_1$s are shown for each in red,
    and the distribution of one type of $F_2$ is shown in purple
    (other types of $F_2$ are not shown).
    Solid lines depict the distance of the $F_2$ to optimum.
    \plri{Should show all types of F2? would be messy.}
}
\end{figure}

@@

If we prefer not to make any assumptions about the shape of the fitness function, we can more conservatively state that hybrid phenotypes diverge from their optimal parents at rates $d(\Phi_{F_1}, \Phi_0) = c_1 \sigma_N^2 T/N_e$ and $d(\Phi_{F_2}, \Phi_0) = \gamma^{-1} \sigma_N \sqrt{T/N_e}$. Whether or not hybrid phenotypic divergence is associated with a high fitness cost, $F_2$ phenotypes should diverge much faster than $F_1$s, with a rate proportional to the square root of $T/N_e$, versus linearly.

@@

%% empirical data on sigma
%The amount and structure of this standing variation is established over long time scales
%by many factors, including
%mutation-selection balance, 
%shifts in the phenotypic optimum,
%and/or spatial variation in the optimum \citep{hansen1996translating}.
%Quantitative genetics models of mutation-selection balance 
%predict precise levels and structure of standing variation \citep{kimura_mutsel,lande_mutsel,lande1981models},
%but it is unclear how well these predictions match reality \citep{johnson_barton}
%and how much they are expected to change over time \citep{arnold_changing_G}.
%However, empirical work allows us to estimate at least the rough magnitude of variation.
%Differences in $A_{ij}$ due to a sequence change are hard to measure,
%but variation in both transcription factor binding site occupancy
%and expression levels (e.g., cis-eQTL) have been measured in various systems.
%However, variation in binding site occupancy may overestimate variation in $\Sys$,
%since it does not capture buffering effects (if for instance only one site of many needs be occupied for transcription to begin),
%and variation in expression levels measures changes in steady-state concentration (our $\kappa_i$) rather than the \emph{rate} of change.
%%Nonetheless,
%%\citet{kasowski2010variation} found differential occupancy in 7.5\%
%%of binding sites of a transcription factor (p65) between human individuals.
%%\citet{verlaan2009targeted} showed that cis-regulatory variation
%%accounts for around 2--6\% of expression variation in human blood-derived primary cells,
%%while \citep{lappalainen2013transcriptome} found that human population variation 
%%explained about 3\% of expression variation.
%%\plr{Get some data from at least one other species in here!}
%%Taken together, this suggests that variation in the entries of $\Sys$
%%may be on the order of 1\% between individuals of a population --
%%doubtless varying substantially between species and between genes.
%%
%Given the empirical data on transcription factor binding and gene expression
%intrapopulation variation \citet{kasowski2010variation, verlaan2009targetedm lappalainen2013transcriptome}, we conservatively estimate that variation in the regulatory coefficients in $\Sys$ may be on the order of 1\% between individuals of a population -- 
%doubtless varying substantially between species and between genes.
%
%%% argument for strength of selection
%%It seems certain that selection in most species is not so strong that intra-population variation
%%is strongly deleterious,
%%so that if $\beta$ is the typical scale on which selection acts,
%%then $\beta > \sigma$.  
%%However,
%%a range of studies \plr{find them} have found evidence for weak stabilizing selection
%%on regulatory SNPs and cis-eQTL.
%For instance, 
%\citet{maria_and_sergey} \plr{(others?)} found evidence that large-effect regulatory mutations are weakly selected against in Drosophila.
%This suggests that 
%the strength of selection on phenotype is sufficient to weakly constrain regulatory variation,
%so that perhaps $\sigma$ and $\beta$ are relatively close.
%This is as would be expected if available variation is held in check by mutation--selection balance
%rather than genetic drift.
%A conservative estimate would be that $\beta = 5 \sigma$;
%taking $\sigma=.01$ as above, this suggests that changes in phenotype of 5\% are sufficient
%to effect a noticeable drop in fitness.
%\plr{BUT $\sigma$ IS VARIATION IN $A$ NOT PHENOTYPE}
%
%We have guessed that within a population, 
%the entries of $\Sys$ vary by around 1\%, 
%at least for networks whose function is strongly constrained.
%
%
\paragraph{Speciation rates under neutrality}
For cases where system structures and $\gamma$ can be inferred,
specific speciation rates can be estimated.
For other situations we offer a back of the envelope calculation.

Consider a population that is distributed on $\mathcal{X}$ such that the magnitude of phenotypically
divergent $\sigma_S$ and neutral $\sigma_N$ diversity are constrained by selection;
$\sigma_S$ directly and $\sigma_N$ indirectly via recombination load.
Two reproductively isolated populations will, on average, be a distance of $2 \sigma_N \sqrt{T/N_e}$.
If we assume that $\sigma_N$ is constrained by a function of $\sigma_S$,
neutral movement along $\mathcal{X}$ pushes hybrids off of the manifold by $\sigma_S$ per $\sigma_N$,
Lastly, if fitness is Gaussian: $\fit_2 = \exp\left(- d(x, x')^2/\beta^2\right)$,
and we assume that the strength of selection is only several multiples of the segregating divergent diversity,
such that $\sigma_S = c \beta$, we can simplify the expression estimating $F_2$ hybrid fitness drops
to be,
  \begin{align}
    \exp\left(-c^2 \frac{T}{N_e} \right)
  \end{align}
%
%$F_2$s between allopatric populations will suffer an average fitness drop relative to their parents of,
%  \begin{align}
%    \exp\left(-\left( \frac{2\sigma_{N}}{\beta}\sqrt{\frac{T}{N_e}} \right)^{2} \right) .
%  \end{align}

%If the strength of selection is only a few multiples $\xi_1$ of the standing phenotypic diversity within a population $\sigma_S$, such that $\beta = \xi_1 \sigma_S$, and recombination load constrains the standing phenotype-neutral diversity $\sigma_N$ to be only a few multiples $\xi_2$ of $\sigma_S$, then $F_2$ fitness drop will be,
%  \begin{align}
%    \exp\left(- \xi^{2} \frac{T}{N_e} \right)
%  \end{align}
%  where $\xi := 2\xi_2/\xi_1$.
  This suggests that, as an upper bound, if a population contains as much phenotypic diversity as allowed by selection, reproductively isolated populations can speciate rapidly -- possibly on timescales shorter than $N_{e}$ generations (It is likely that, on average, $c \leq 2/3$).
%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%

%    \subsection*{Metabolic Network}
%  \begin{align*}
%    \dot{x}(t) &= \begin{bmatrix} 0 & 1 & 0 & -1 \\ 0 & 0 & -1 & -1 \\ -1 & 1 & 0 & 0 \\ 0 & 0 & 0 & 0 \end{bmatrix} \begin{bmatrix} \text{GAL3} \\ \text{GAL4} \\ \text{GAL80} \\ \text{MIG1} \end{bmatrix}(t) \\ &+ \begin{bmatrix} 1 & 0 & 0 \\ 0 & 0 & 1 \\ 0 & 0 & 0 \\ 0 & 1 & 1 \end{bmatrix} \begin{bmatrix} \text{galactose} \\ \text{glucose} \\ \text{BTM} \end{bmatrix}(t) \\
%        y(t) &= \begin{bmatrix} 0 & 1 & 0 & -1 \end{bmatrix} \vec{x}(t)
%    \end{align*}
%    \begin{align*}
%      H(z) &= \begin{bmatrix} 1 \\ z^{2} + z + 1 \\ z + 1 \end{bmatrix} \frac{1}{z^{3} + z -1}
%    \end{align*}
%
%    \begin{align*}
%      CV &= C \\
%      VB &= B \\
%      V :&= \begin{bmatrix} 1 & 0 & a & 0 \\ 0 & 1 & 0 & 0 \\ 0 & 0 & b & 0 \\ 0 & 0 & 0 & 1 \end{bmatrix} , V^{-1} = \begin{bmatrix} 1 & 0 & \frac{-a}{b} & 0 \\ 0 & 1 & 0 & 0 \\ 0 & 0 & \frac{1}{b} & 0 \\ 0 & 0 & 0 & 1\end{bmatrix} \\
%        A(a,b) :&= VAV^{-1} = \begin{bmatrix} -a & a+1 & \frac{a^{2}}{b} & -1 \\ 0 & 0 & \frac{-1}{b} & -1 \\ -b & b & a & 0 \\ 0 & 0 & 0 & 0 \end{bmatrix}
%    \end{align*}
%    The GAL regulatory network in \emph{S. cerevisiae} is modelled here. The activating transcription factor GAL4, which is regulated by the presence of galactose and two other TFs, binds to the promoter region of the GAL regulon -- a cluster of genes regulated by the same promoter encoding 3 enzymes (GAL1, GAL7, and GAL10), for the metabolism of the sugar galactose. The repressing transcription factor MIG1, which is activated by the presence of glucose and other proteins omitted here, also binds to the regulon, preventing the expression of the galactose metabolizing enzymes. This network is one of the most studied gene regulatory networks in yeast, and experiments have already demonstrated significant transcriptional variation among different species and genuses of yeast.
%
%    In the present model, the $A$ matrix encodes the interactions among GAL3, GAL4, GAL80, and MIG1. $B$ interprets the input: the quantity of glucose, galactose, and the basal transcription (BTM) rates impact of TF concentrations. $C$ represents the promoter region of the GAL regulon and is the sum of the influence of GAL4 (activating) and MIG1 (repressing).
%
%    $A(a,b)$ represent alternative network structures that produce identical outputs, with identical input and output transformations -- meaning the regulatory contributions of the sugars and the basasl machinery, as well as the promoter of the GAL regulon are held constant. These interactions may be realized in nature, if the regulatory differences predicted by $A(a,b)$ are biologically/mutationally realizable. Mutations influencing the interactions between GAL4 and GAL80 have been experimentally demonstrated \citep{li2010alterations}.
%
%    \subsection*{Gap Gene Network}

%%%%%%%%%%%%%%%%%%%%%
\section*{Discussion}
%
%\jss{Discussion guidelines:
%\begin{itemize}
%    \item Why is this important/useful?
%    \item What are the assumptions and shortcomings of the research?
%    \item Compare to other studies in the literature. 
%    \item Future directions. 
%    \item Wild speculations?
%    \item Conclusion and overall impact. 
%\end{itemize}}
%
The complexity of biological systems has limited our understanding of their function and evolution. Above we outline an approach, a first step, towards untangling this complexity in reference to function and evolution. This methodology borrows successfully applied tools from engineering and aims to synthesize these with the concepts and tools of molecular and evolutionary biology.

@@

%We also suggest that gene networks may not always use their components parsimoniously 
%as network size tends to ratchet up in the absence of strong selection against extra parts. 
%\plr{probably shouldn't claim this is "unexplored"} \js{I meant unexplored within this paper.Changed wording. Does that work?}
%Although we leave this question unexplored, this phenomena may lead to insights on evolvability and developmental innovation. 
Lastly, we show that hybrid gene networks break down as function of genetic distance, and may, in part, explain broad patterns of reproductive isolation among diverse phyla \citep{roux2016shedding}.

>>>>

\documentclass{article}
\usepackage{amsmath, amssymb, color, xcolor, amsthm}
\usepackage{graphicx, wrapfig, float, caption, dsfont, bbm, xfrac}
\usepackage{fullpage}
\usepackage[backref=page, hidelinks, colorlinks=true, citecolor=blue!60!black!100]{hyperref}
\usepackage{tikz}
\usetikzlibrary{arrows.meta, shapes}
\usepackage{caption, subcaption}
\usepackage{natbib} % gives us \citet: Author (year) and \citep: (Author; year)
\usepackage{authblk}

@@

\begin{figure}[H]
\centering
\includegraphics{figures/conceptual_fig}
\caption{
    \label{fig:conceptual_fig}
    A conceptual figure of the fitness consequences of hybridization:
    axes represent system coefficients (i.e., entries of $A$);
    the line of optimal system coefficients is down in black;
    dotted lines give phenotypic distances to the optimum.
    Two pairs of parental populations are shown in black, along the optimum;
    a hypothetical population of $F_1$s are shown for each in red,
    and the distribution of one type of $F_2$ is shown in purple
    (other types of $F_2$ are not shown).
    Solid lines depict the distance of the $F_2$ to optimum.
    \plri{Should show all types of F2? would be messy.}
}
\end{figure}

@@

Note that this follows directly from the fact that $F_1$ and $F_2$ hybrid phenotypes diverge from $F_0$ phenotypes, linearly and inverse-quadratically, with respect to time (more precisely $T/N_e$). As such, removing assumptions about the form of the fitness function, this model still predicts hybrid $F_2$ phenotypes to diverge much faster than $F_1$s, initially.

@@

%% empirical data on sigma
%The amount and structure of this standing variation is established over long time scales
%by many factors, including
%mutation-selection balance, 
%shifts in the phenotypic optimum,
%and/or spatial variation in the optimum \citep{hansen1996translating}.
%Quantitative genetics models of mutation-selection balance 
%predict precise levels and structure of standing variation \citep{kimura_mutsel,lande_mutsel,lande1981models},
%but it is unclear how well these predictions match reality \citep{johnson_barton}
%and how much they are expected to change over time \citep{arnold_changing_G}.
%However, empirical work allows us to estimate at least the rough magnitude of variation.
%Differences in $A_{ij}$ due to a sequence change are hard to measure,
%but variation in both transcription factor binding site occupancy
%and expression levels (e.g., cis-eQTL) have been measured in various systems.
%However, variation in binding site occupancy may overestimate variation in $\Sys$,
%since it does not capture buffering effects (if for instance only one site of many needs be occupied for transcription to begin),
%and variation in expression levels measures changes in steady-state concentration (our $\kappa_i$) rather than the \emph{rate} of change.
%%Nonetheless,
%%\citet{kasowski2010variation} found differential occupancy in 7.5\%
%%of binding sites of a transcription factor (p65) between human individuals.
%%\citet{verlaan2009targeted} showed that cis-regulatory variation
%%accounts for around 2--6\% of expression variation in human blood-derived primary cells,
%%while \citep{lappalainen2013transcriptome} found that human population variation 
%%explained about 3\% of expression variation.
%%\plr{Get some data from at least one other species in here!}
%%Taken together, this suggests that variation in the entries of $\Sys$
%%may be on the order of 1\% between individuals of a population --
%%doubtless varying substantially between species and between genes.
%%
%Given the empirical data on transcription factor binding and gene expression
%intrapopulation variation \citet{kasowski2010variation, verlaan2009targetedm lappalainen2013transcriptome}, we conservatively estimate that variation in the regulatory coefficients in $\Sys$ may be on the order of 1\% between individuals of a population -- 
%doubtless varying substantially between species and between genes.
%%% argument for strength of selection
%%It seems certain that selection in most species is not so strong that intra-population variation
%%is strongly deleterious,
%%so that if $\beta$ is the typical scale on which selection acts,
%%then $\beta > \sigma$.  
%%However,
%%a range of studies \plr{find them} have found evidence for weak stabilizing selection
%%on regulatory SNPs and cis-eQTL.
%For instance, 
%\citet{maria_and_sergey} \plr{(others?)} found evidence that large-effect regulatory mutations are weakly selected against in Drosophila.
%This suggests that 
%the strength of selection on phenotype is sufficient to weakly constrain regulatory variation,
%so that perhaps $\sigma$ and $\beta$ are relatively close.
%This is as would be expected if available variation is held in check by mutation--selection balance
%rather than genetic drift.
%A conservative estimate would be that $\beta = 5 \sigma$;
%taking $\sigma=.01$ as above, this suggests that changes in phenotype of 5\% are sufficient
%to effect a noticeable drop in fitness.
%\plr{BUT $\sigma$ IS VARIATION IN $A$ NOT PHENOTYPE}
%We have guessed that within a population, 
%the entries of $\Sys$ vary by around 1\%, 
%at least for networks whose function is strongly constrained.
\paragraph{Speciation rates under neutrality}
%For cases where system structures and $\gamma$ can be inferred,
%specific speciation rates can be estimated.
%For other situations we offer a back of the envelope calculation.
%Consider a population that is distributed on $\mathcal{X}$ such that the magnitude of phenotypically
%divergent $\sigma_S$ and neutral $\sigma_N$ diversity are constrained by selection;
%$\sigma_S$ directly, such that $\sigma_S = \gamma$, and $\sigma_N$ indirectly via recombination load.
%Two reproductively isolated populations will, on average, be a distance of $2 \sigma_N \sqrt{T/N_e}$.
%If we assume that $\sigma_N$ is constrained by a function of $\sigma_S$,
%neutral movement along $\mathcal{X}$ pushes hybrids off of the manifold by $\sigma_S$ per $\sigma_N$,
%Lastly, if fitness is Gaussian: $\fit_2 = \exp\left(- d(x, x')^2/\beta^2\right)$,
%and we assume that the strength of selection is only several multiples of the segregating divergent diversity,
%such that $\sigma_S = c \beta$, we can simplify the expression estimating $F_2$ hybrid fitness drops
%to be,
%$F_2$s between allopatric populations will suffer an average fitness drop relative to their parents of,
%    \exp\left(-\left( \frac{2\sigma_{N}}{\beta}\sqrt{\frac{T}{N_e}} \right)^{2} \right) .

%If the strength of selection is only a few multiples $\xi_1$ of the standing phenotypic diversity within a population $\sigma_S$, such that $\beta = \xi_1 \sigma_S$, and recombination load constrains the standing phenotype-neutral diversity $\sigma_N$ to be only a few multiples $\xi_2$ of $\sigma_S$, then $F_2$ fitness drop will be,
%  \begin{align}
%    \exp\left(- \xi^{2} \frac{T}{N_e} \right)
%  \end{align}
%  where $\xi := 2\xi_2/\xi_1$.
  If we assume that perpendicular variation segregating within a population is constrained by selection such that $\sigma_S/\gamma = \xi$, and that neutral variation is on the order of perpendicular variation, such that $\sigma_N = \sigma_S$, then the typical distance between two individuals drawn from separate allopatric populations will be $z = 2 \sigma_N \sqrt{T/N_e}$. We will expect a drop in hybrid fitness to be,
  \begin{align}
  \fit_2/\fit_0 = \exp \left(- 4\xi^2 \frac{T}{N_e} \right) .
  \end{align}
  Hybrid fitness depression will be significant when $z > \gamma$. 
  If $\xi$ is in $[\sfrac{1}{5},1]$, speciation will occur on the order of $N_e$ generations, and possibly much faster. Note, however, that if intrapopulation variation is very small such that $\sigma_S \ll \gamma$, speciation rates will be slower.
%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%
\section*{Discussion}
%
%\jss{Discussion guidelines:
%\begin{itemize}
%    \item Why is this important/useful?
%    \item What are the assumptions and shortcomings of the research?
%    \item Compare to other studies in the literature. 
%    \item Future directions. 
%    \item Wild speculations?
%    \item Conclusion and overall impact. 
%\end{itemize}}
%
The complexity of biological systems has limited our understanding of their function and evolution. Above we outline an approach, a first step, towards untangling this complexity in reference to function and evolution. This methodology borrows successfully applied tools from engineering and aims to synthesize these with the concepts and tools of molecular and evolutionary biology.

@@

%We also suggest that gene networks may not always use their components parsimoniously 
%as network size tends to ratchet up in the absence of strong selection against extra parts. 
%\plr{probably shouldn't claim this is "unexplored"} \js{I meant unexplored within this paper.Changed wording. Does that work?}
%Although we leave this question unexplored, this phenomena may lead to insights on evolvability and developmental innovation. 
Lastly, we show that hybrid gene networks, under neutral processes, break down as function of genetic distance, and may, in part, explain broad patterns of reproductive isolation among diverse phyla \citep{roux2016shedding}.
